Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-05-07 paper/Paper.Rnw
moved some more stuff to the SI for length reasons

\noindent \textbf{Running head:} Asymmetric relationships and coexistence \\
\noindent \textbf{Key words:} asymmetric tail association, biodiversity, coexistence, competition, copula, modern coexistence theory, niche differentiation, storage effects \\
\noindent \textbf{Type of article:} Letter\\
\noindent \textbf{Words in Abstract:} 150; \textbf{Words in main text:} 5518; \textbf{Words in text boxes:} 0\\
\noindent \textbf{Number of references:} 39; \textbf{Number of figures:} 5; \textbf{Number of tables:} 1; \textbf{Number of text boxes:} 0\\
\noindent \textbf{Corresponding author:}
Daniel C. Reuman,
Department of Ecology and Evolutionary Biology,
University of Kansas,
2101 Constant Ave,
Lawrence, KS 66047;
Phone: 626-560-7084 \\
\noindent \textbf{Author emails:} pimsupaa@uoregon.edu, reuman@ku.edu\\
\noindent \textbf{Statement of Authorship:} The authors jointly carried out all parts of the study.\\
\noindent \textbf{Data accessibility statement:} All codes associated with this study will be released
in a Zenodo repository prior to publication.

@@

%E and vital rates
Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', 
in fact it is often instead a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to commonly measured environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is probably better conceptualized as a species response.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
So the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which matter for coexistence.
Vital rates are measured directly in demographic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not been done.

%Differences with the original, analytic version of the MCT
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to expand the mathematics of the original MCT to consider ATAs.

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, a decomposition
of the GWR of $i$ for which some terms were based on rendering one or 
more of $X$, $Y$, and $Z$ independent of the others. 
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}). See SI section \ref{SI-sect:threeway} for details.

@@

\begin{figure}
%\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{ATA effects and ATA rescue in the diatom system. 
The contribution of ATAs to the GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition (a-c). Panels a-c show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dotted lines. 
ATA contributions ($\Delta_i^{[EC]}$)
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to coexistence in this system. In some instances, the GWR line falls 
between the ATA line
and the horizontal axis, indicating that, for those parameter values, ATA rescue occurs. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
Panels d-f show bivariate plots (Methods) for which the third parameter was 
fixed at the values used by \cite{Descamps_2005}. ATAs are again often an important factor helping
determine coexistence or exclusion, and ATA rescue happened for an appreciable fraction of the
explored parameter space.
For the parameters we considered, GWR of the other diatom species in the system, 
\emph{C. pseudostelligera}, was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.
}  \label{fig:fig5}
\end{figure}

>>>>

\noindent \textbf{Running head:} Asymmetric relationships and coexistence \\
\noindent \textbf{Key words:} asymmetric tail association, biodiversity, coexistence, competition, copula, modern coexistence theory, niche differentiation, storage effects \\
\noindent \textbf{Type of article:} Letter\\
\noindent \textbf{Words in Abstract:} 150; \textbf{Words in main text:} 5013; \textbf{Words in text boxes:} 0\\
\noindent \textbf{Number of references:} 39; \textbf{Number of figures:} 4; \textbf{Number of tables:} 1; \textbf{Number of text boxes:} 0\\
\noindent \textbf{Corresponding author:}
Daniel C. Reuman,
Department of Ecology and Evolutionary Biology,
University of Kansas,
2101 Constant Ave,
Lawrence, KS 66047;
Phone: 626-560-7084 \\
\noindent \textbf{Author emails:} pimsupaa@uoregon.edu, reuman@ku.edu\\
\noindent \textbf{Statement of Authorship:} The authors jointly carried out all parts of the study.\\
\noindent \textbf{Data accessibility statement:} All codes associated with this study will be released
in a Zenodo repository prior to publication.

@@

%Differences with the original, analytic version of the MCT
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to expand the mathematics of the original MCT to consider ATAs.
Additional discussion points are in SI sections \ref{SI-sect:vital_rates}-\ref{SI-sect:threeway}.

@@

%***DAN: Jasmin, this currently pulls in a pdf you sent me, stored temporarily on my drive in the indicated 
%location. Once the code produces fig 5 on a normal run, we'll have to change the below to suck that
%version of the figure up instead of this one.
\begin{figure}
\includegraphics[width=\textwidth]{fig5_new.pdf}
\caption{See next page for caption.}  \label{fig:fig5}
\end{figure}

\singlespacing

\noindent Figure 4. ATA effects and ATA rescue in the diatom system. 
The contribution of ATAs to the GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition (a-c). Panels a-c show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dotted lines. 
ATA contributions ($\Delta_i^{[EC]}$)
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to coexistence in this system. In some instances, the GWR line falls 
between the ATA line
and the horizontal axis, indicating that, for those parameter values, ATA rescue occurs. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
Panels d-f show bivariate plots (Methods) for which the third parameter was 
fixed at the values used by \cite{Descamps_2005}. ATAs are again often an important factor helping
determine coexistence or exclusion, and ATA rescue happened for an appreciable fraction of the
explored parameter space.
For the parameters we considered, GWR of the other diatom species in the system, 
\emph{C. pseudostelligera}, was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.

### 2023-05-07 paper/SupMat.Rnw
moved some more stuff to the SI for length reasons

@@

>>>>

\section{   Asymmetric associations between population vital rates}\label{sect:vital_rates}

Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', 
in fact it is often instead a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to commonly measured environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is probably better conceptualized as a species response.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
So the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which matter for coexistence.
Vital rates are measured directly in demographic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not been done.

@@

\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, a decomposition
of the GWR of $i$ for which some terms were based on rendering one or 
more of $X$, $Y$, and $Z$ independent of the others. 
We comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}). Details follow.
